Implicit yaddayadda

### 2011-10-26 papers/PLDI2012/implicitization.tex
explications

We conducted extensive experiments on real-life programs in order to compare the
different techniques. We also compared the precision of the analysis depending
on the choice of the abstract domain.

In order to compare the various techniques, we compare by inclusion the obtained
abstract values at each control points in $P_R$.

In this section, our experimental results come from the analysis of various
open-source projects listed in Table \ref{fig:projects}, written in C, C++ and
Fortran.
\begin{table}[!h]
	\centering
\begin{tabular}{|l|r|r|} \hline
	\multicolumn{1}{|c|}{Name} &
        \multicolumn{1}{c|}{kLOC} &
        \multicolumn{1}{c|}{$|P_R|$} \\ \hline
	a2ps & 55 & 2012\\
	gawk & 59 & 902\\ 
	gnuchess & 38 & 1222\\ 
	gnugo & 83 & 2801\\
	grep & 35 & 820\\
	gzip & 27 & 494\\
	lapack/blas & 954 & 16422\\
	make & 34 & 993\\ 
	sed & 23 & 292\\
	tar & 73 & 1712\\
	\hline
\end{tabular}
\caption{List of analyzed open-source projects, with their respective number of
lines of code, and their number of control points in $P_R$}
\label{fig:projects}
\end{table}

@@

Table \ref{tab:techniques} gives the results of the comparison between the
different analysis techniques, applied to the benchmarks described in Figure
\ref{fig:projects}. These results are also summarized by Figure \ref{fig:techniques}.

\emph{Guided Static Analysis} from \citet{DBLP:conf/sas/GopanR07} improves the
result of the classical Abstract Interpretation in $2.37\%$ of the control points
in $P_R$.
\emph{Path-focused} technique from \citet{Monniaux_Gonnord_SAS11} gives
statistically better results, and finds better invariants than \emph{Guided
Static Analysis} in $7.28\%$ of the cases. However, it also looses precision in an
important number ($2.93\%$) of control points.
Our combined technique gives the most promising results, since it is
statistically more precise than the other techniques.

@@

While experiencing with techniques that use SMT-solving, we encountered some
limitations due to non-linear arithmetic in the analyzed programs. Indeed, 
the SMT-solver is not able to decide the satisfiability of some SMT-formulae
expressing the semantics of non-linear programs. 
In this case, we skipped the function for which the SMT-solver returned the
``unknown'' result.
This limitation occurred very rarely in our experiments, except for the analysis
of \emph{Lapack/Blas}, where 798 over the 1602 functions have been skipped.
In such cases, such a technique should make the SMT-formula simpler to solve by using
for instance \emph{linearization}. However, such transformations are not
implemented in our analyzer.

>>>>

We conducted extensive experiments on real-life programs in order to compare the
different techniques, mostly on open-source projects (Fig.~\ref{fig:projects}) written in C, C++ and Fortran.

\begin{table}[!h]
	\centering
\begin{tabular}{|l|r|r|} \hline
	\multicolumn{1}{|c|}{Name} &
        \multicolumn{1}{c|}{kLOC} &
        \multicolumn{1}{c|}{$|P_R|$} \\ \hline
	a2ps & 55 & 2012\\
	gawk & 59 & 902\\ 
	gnuchess & 38 & 1222\\ 
	gnugo & 83 & 2801\\
	grep & 35 & 820\\
	gzip & 27 & 494\\
	lapack/blas & 954 & 16422\\
	make & 34 & 993\\ 
	sed & 23 & 292\\
	tar & 73 & 1712\\
	\hline
\end{tabular}
\caption{List of analyzed open-source projects, with their respective number of
lines of code, and their number of control points in $P_R$}
\label{fig:projects}
\end{table}

@@

We compared the precision of the different techniques described in this article as follows:
\begin{itemize}
\item For each program, we distinguished a set $P_R = P_W$ of suitable widening points by a simple algorithm: for each procedure, we compute the strongly connected components of its control-flow graph using Tarjan's algorithm; the targets of the back-edges of the depth-first search are added to~$P_R$. Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles.
(More sophisticated techniques are discussed in e.g. \citet{BourdonclePhd}).

\item For each program and each pair $(T_1,T_2)$ of analysis techniques, we list the proportion of control points in $P_R$ where $T_1$ (resp.~$T_2$) gives a strictly stronger invariant, denoted by $\subsetneq$ (resp. $\supsetneq$), and the proportion of control points where the invariants given by $T_1$ and $T_2$ are uncomparable for the inclusion ordering (the remainder of the control points are thus those for which both techniques give the same invariant).
\end{itemize}

Our analysis tool currently only operates over scalar variable and cannot directly cope with arrays or memory accessed through pointers. We therefore run it after the ``memory to registers'' (\texttt{mem2reg}) optimization phase in LLVM, which lifts most memory accesses to scalar variables.
The remaining memory accesses are treated as undefined values, as are the return values of function calls. For better precision, we also apply function inlining to the program.

The analyses are performed using convex polyhedra as abstract domain. Variables defined by linear expressions over other variables (e.g. $y=x+1$) are not retained as dimensions in the polyhedra; therefore the only variables appearing in the polyhedra are those defined by $\phi$ operators (join points) in the SSA representation, and those introduced by constructs we cannot model as linear operations, such as loads from memory, return values from function calls, or bitwise arithmetic.

% TODO: avec quel domaine abstrait?
% TODO: discuter du big step SSA?
% TODO: cite SSA

Let us briefly comment the results given in more details in Table~\ref{tab:techniques} and Figure~\ref{fig:techniques}.
\emph{Guided Static Analysis} from \citet{DBLP:conf/sas/GopanR07} improves the
result of the classical Abstract Interpretation in $2.37\%$ of the control points
in $P_R$.
\emph{Path-focusing} from \citet{Monniaux_Gonnord_SAS11} gives
statistically better results, and finds better invariants than \emph{Guided
Static Analysis} in $7.28\%$ of the cases. However, it also loses precision in an
important number ($2.93\%$) of control points.
Finally, our combined technique gives the most promising results, since it is
statistically more precise than the other techniques.

@@

While experiencing with techniques that use SMT-solving, we encountered some
limitations due to non-linear arithmetic in the analyzed programs. Indeed, 
the SMT-solver is not able to decide the satisfiability of some SMT-formulae
expressing the semantics of non-linear programs. 
In this case, we skipped the functions for which the SMT-solver returned the
``unknown'' result.
This limitation occurred very rarely in our experiments, except for the analysis
of \emph{Lapack/Blas}, where 798 over the 1602 functions have been skipped.
\emph{Lapack/Blas} implements matrix computations, which use floating-point multiplications.

In cases where the formula is expressed in too rich a logic for the SMT-solver to deal with, a number of workarounds are possible:
\begin{itemize}
\item \emph{Linearization}, as per \citet{DBLP:conf/vmcai/Mine06}, which overapproximates nonlinear semantics by linear semantics.
\item Replacing the results of nonlinear operations by ``unknown''.
\end{itemize}
However, such transformations are not yet implemented in our tool.
% TODO: c'est si dur de remplacer * non lineaire par undef?
